### 2015-07-16 paper.tex
putting paper in repo

\documentclass[12pt]{article}
\usepackage{amsmath,amsthm,amssymb}
\usepackage{graphicx}
\usepackage{cite}
\pagestyle{myheadings}
\author{Joseph McDonald}

\textwidth=7in
\textheight= 9in

\topmargin=-0.5in
\oddsidemargin=-0.25in

\newcommand{\xbar}{\bar{x}}
\newcommand{\V}[1]{\boldsymbol{#1}}
\newcommand{\E}[0]{\mathbb{E}}
\title{Parallelizing AdaBoost and Decision Trees}

\begin{document}
\maketitle

\section{Introduction}
Binary and multiclass classification is a fundamental problem in machine
learning and is commonly one of the first concepts to be studied in an
introductory course, the goal being to accurately predict the label of a given
observations using the data associated with the observation. In the context of
binary classification, a labeled observation (or sample point) is a coordinate
pair $(x, y)$ where $x\in\mathbb{R}^d$ is a vector of data associated with the
observation (feature vector) and $y\in\{1, -1\}$ is the corresponding label of
the observation. This classification, and even prediction in general, commonly
splits up into two different scenarios, supervised and unsupervised learning.
Supervised learning involves taking a set of labeled observations (training
set) and using it to construct a prediction function, one that can take a new
feature vector and predict the corresponding label. There are many methods used
for classification in supervised learning, including support vector machines,
decision trees, online algorithms, and many others.

A weak learning algorithm, or weak learner, is an algorithm that, given a set of training data, yields a predictor with performance of at least 50\% percent accuracy, or just better than random guessing. The prediction functions that a weak learner returns are called base classifiers. There was an important question asking whether the base classifiers from a weak learner could somehow be joined together to produce a strong learner, or an algorithm that gives a prediction function of arbitrary accuracy.

Boosting is a category of methods that addresses this question, which Freund and Schapire
successfully answered with their algorithm AdaBoost. AdaBoost, which stands for adaptive boosting, builds
a more accurate predictor through successive rounds of training a weak learner
to the data and subsequently re-weighting the training samples \cite{no1}.

\begin{thebibliography}
\ref[no1] Rob Schapire

\end{thebibliography}
\end{document}

### 2015-07-20 paper.tex
adding paper

@@
 \begin{document}
+
+%\begin{abstract}
+%I present a parallelized construction of decision trees to be used in AdaBoost
+%\end{abstract}
+
 \maketitle
@@
 \section{Introduction}
-Binary and multiclass classification is a fundamental problem in machine
-learning and is commonly one of the first concepts to be studied in an
-introductory course, the goal being to accurately predict the label of a given
-observations using the data associated with the observation. In the context of
-binary classification, a labeled observation (or sample point) is a coordinate
-pair $(x, y)$ where $x\in\mathbb{R}^d$ is a vector of data associated with the
-observation (feature vector) and $y\in\{1, -1\}$ is the corresponding label of
-the observation. This classification, and even prediction in general, commonly
-splits up into two different scenarios, supervised and unsupervised learning.
-Supervised learning involves taking a set of labeled observations (training
-set) and using it to construct a prediction function, one that can take a new
-feature vector and predict the corresponding label. There are many methods used
-for classification in supervised learning, including support vector machines,
-decision trees, online algorithms, and many others.
+Binary and multiclass classification is a fundamental problem in machine learning and is commonly one of the first concepts to be studied in an introductory course.
+In the context of binary classification, a labeled observation is a coordinate pair $(x, y)$ where $x\in\mathbb{R}^d$ is a vector of data associated with the observation (feature vector) and $y\in\{1, -1\}$ is the corresponding label of the observation.
+For example, $x_i$ might be a point in $\mathbb{R}^2$ representing the height and weight of the $i$th individual in a data set while the value of $y_i$ may correspond to the person's gender.
+The goal in classification is to accurately predict the label of a given observation using its feature vector.
+In this setting, the scenario of supervised learning involves using a training set of labeled observations $(x_i, y_i)$ to construct a prediction function $h$, one that takes a feature vector $\hat{x}$ and gives a prediction on the corresponding label $h(\hat{x})=\hat{y}$.
+It is common to use a separate set of test data to check the accuracy of the predictor, data that hasn't been used for training.
+There are many methods used to obtain a predictor for classification in supervised learning, and the class of ensemble methods concerns combining several predictors to obtain one.
 
-A weak learning algorithm, or weak learner, is an algorithm that, given a set of training data, yields a predictor with performance of at least 50\% percent accuracy, or just better than random guessing.
-The prediction functions that a weak learner returns are called base classifiers.
-There was an important question asking whether the base classifiers from a weak learner could somehow be joined together to produce a strong learner, or an algorithm that gives a prediction function of arbitrary accuracy.
+Suppose now there exists an algorithm that, for any set of training data, is guaranteed to give you a predictor that has at least 50\% accuracy.
+Such an algorithm is called a weak learning algorithm since it yields a predictor that performs marginally better than guessing randomly.
+These prediction functions that a weak learning algorithm returns are called base classifiers, and a natural question arises about whether we can use a combination of weak learners to get a stronger one with much better accuracy and what this sort of ensemble method would look like.
+Ensemble methods like this are called boosting, and the algorithm AdaBoost introduced in \cite{no1} accomplishes this through successive rounds of training a weak learner to the data and subsequently re-weighting the training samples \cite{no1}.
 
-Boosting is a category of methods that addresses this question, which Freund and Schapire
-successfully answered with their algorithm AdaBoost. AdaBoost, which stands for adaptive boosting, builds
-a more accurate predictor through successive rounds of training a weak learner
-to the data and subsequently re-weighting the training samples \cite{no1}.
+The goal of this project is to implement a parallelized version of a commonly used weak learner, decision trees, to improve the running time of AdaBoost.
+We give a short overview of AdaBoost and decision trees in the next section, describe the implementation of the decision trees, the data sets used and the results t
+
+\section{AdaBoost}
+The beauty of AdaBoost is in using a relatively inaccurate weak learning algorithm to produce a much more accurate prediction function.
+As mentioned, in each iteration of AdaBoost the weak learner selects a base classifier $h: \mathbb{R}^n\rightarrow\{1,-1\}$ from a set of functions $H$.
+The weak learner chooses in particular the base classifier  We give the algorithm below.\\
+\\
+{\sc AdaBoost}: $S = \{(x_i,y_i):i = 1\ldots n\}$
+\begin{enumerate}
+\itemsep1pt \parskip0pt \parsep0pt
+\item {\bf for $i=1$ to $n$}
+\item \quad $D_1(i) = \frac{1}{n}$
+\item {\bf for $t=1$ to $T$}
+\item \quad $h_t$ = {\sc WeakLearner}$(S\sim D_t)$ with weighted error $\epsilon_t$
+\item \quad $\alpha_t = \frac{1}{2}\log(\frac{1-\epsilon_t}{\epsilon_t})$
+\item \quad $Z_t = 2[\epsilon_t(1-\epsilon_t)]^{1/2}$ (normalization factor)
+\item \quad {\bf for $i = 1$ to $n$}
+\item \quad \quad $D_{t+1}(i) = \frac{D_t(i)\exp(-\alpha_t y_i h_t(x_i))}{Z_t}$
+\item $g = \sum_{t=1}{T} \alpha_t h_t$
+\item {\bf return} $\mbox{sign}(g)$
+\end{enumerate}
+
+\section{Decision Trees}
 

### 2015-08-01 paper.tex
paper changes

@@
 \section{Decision Trees}
+Decision trees are a well-known method useful for classification which partitions the feature space into rectangular regions and gives each region the most common label among the training set points it contains.
+Each partition of the feature space is made along coordinate axes and done in a heirarchical way.
+Starting with the entire space, a line parallel to an axis divides it into two regions and then the same is done for these resulting regions.
+This process is repeated many times until the space is partitioned into several rectangles, like the example in Figure 1 (\cite{HTF}).
+\begin{figure}
+\includegraphic{dectreerect}
+\end{figure}
+
+ In that depiction, the lines drawn at $t_i$ represent the $i$th division of the space.
+Since each region is divided in two, this gives rise to a binary tree where each node represents a line drawn to partition the region existing before that node.
+The tree corresponding to Figure 1 is given in Figure 2 (\cite{HTF}).
+
+ It is done in a heirarchical way that The feature space is partitioned along coordinate axes, which gives rise to decision.
+The most common way to partition the feature space is along coordinate axes.
+
+%\maketitle
+%
+%\section{Introduction}
+%Binary and multiclass classification is a fundamental problem in machine
+%learning and is commonly one of the first concepts to be studied in an
+%introductory course, the goal being to accurately predict the label of a given
+%observations using the data associated with the observation. In the context of
+%binary classification, a labeled observation (or sample point) is a coordinate
+%pair $(x, y)$ where $x\in\mathbb{R}^d$ is a vector of data associated with the
+%observation (feature vector) and $y\in\{1, -1\}$ is the corresponding label of
+%the observation. This classification, and even prediction in general, commonly
+%splits up into two different scenarios, supervised and unsupervised learning.
+%Supervised learning involves taking a set of labeled observations (training
+%set) and using it to construct a prediction function, one that can take a new
+%feature vector and predict the corresponding label. There are many methods used
+%for classification in supervised learning, including support vector machines,
+%decision trees, online algorithms, and many others.
+%
+%A weak learning algorithm, or weak learner, is an algorithm that, given a set of training data, yields a predictor with performance of at least 50\% percent accuracy, or just better than random guessing.
+The prediction functions that a weak learner returns are called base classifiers.
+There was an important question asking whether the base classifiers from a weak learner could somehow be joined together to produce a strong learner, or an algorithm that gives a prediction function of arbitrary accuracy.
+%
+%Boosting is a category of methods that addresses this question, which Freund and Schapire
+%successfully answered with their algorithm AdaBoost. AdaBoost, which stands for adaptive boosting, builds
+%a more accurate predictor through successive rounds of training a weak learner
+%to the data and subsequently re-weighting the training samples \cite{no1}.
 
 \begin{thebibliography}
-\ref[no1] Rob Schapire
+\ref[no1] Yoav Freund Rob Schapire
+
+\ref[no2] Hastie Friedman Tibshirani
+
+\ref[no3] Mohri
+
+\ref[no4] Wiki
 

### 2015-08-07 paper.tex
paper changes

@@
 \section{Decision Trees}
-Decision trees are a well-known method useful for classification which partitions the feature space into rectangular regions and gives each region the most common label among the training set points it contains.
-Each partition of the feature space is made along coordinate axes and done in a heirarchical way.
-Starting with the entire space, a line parallel to an axis divides it into two regions and then the same is done for these resulting regions.
-This process is repeated many times until the space is partitioned into several rectangles, like the example in Figure 1 (\cite{HTF}).
-\begin{figure}
-\includegraphic{dectreerect}
+Decision trees are a well-known method useful for classification which
+partitions the feature space into rectangular regions and gives each region the
+most common label among the training set points it contains. Each partition of
+the feature space is made along coordinate axes and done in a heirarchical way.
+Starting with the entire space, a line parallel to an axis divides it into two
+regions and then the same is done for these resulting regions. This process is
+repeated many times until the space is partitioned into several rectangles,
+like the example in Figure 1 (\cite{HTF}).
+Each split is made to separate points of the two classes distinct classes into different regions as best as possible.
+Thus the two resulting rectangles will have a greater {\it purity} in comparison to the parent node.
+There are a .
+
+\begin{figure}[b!]
+\centering
+\begin{minipage}{0.45\textwidth}
+\centering
+\includegraphics[width=2.5in]{partition}
+\caption{space partitioned by a decision tree}
+\end{minipage}\hfill
+\begin{minipage}{0.45\textwidth}
+\centering
+\includegraphics[width=2.5in]{dectree}
+\caption{decision tree}
+\end{minipage}
 \end{figure}
@@
 The tree corresponding to Figure 1 is given in Figure 2 (\cite{HTF}).
-
- It is done in a heirarchical way that The feature space is partitioned along coordinate axes, which gives rise to decision.
+As depicted in the two figures, the splitting of each node is done in a heirarchical way so that .
+The feature space is partitioned along coordinate axes, which gives rise to decision.
 The most common way to partition the feature space is along coordinate axes.
@@
 
-\begin{thebibliography}
-\ref[no1] Yoav Freund Rob Schapire
+%\begin{thebibliography}
+%\bibitem{FS} Yoav Freund Rob Schapire
 
-\ref[no2] Hastie Friedman Tibshirani
+%\bibitem{HTF} Hastie Friedman Tibshirani
 
-\ref[no3] Mohri
+%\bibitem{Mohri} Mohri
 
-\ref[no4] Wiki
-
-\end{thebibliography}
+%\bibitem{no4} Wiki
+%\end{thebibliography}
 \end{document}

### 2015-08-18 paper.tex
changes to paper

@@
 
-\section{Introduction}
-Binary and multiclass classification is a fundamental problem in machine learning and is commonly one of the first concepts to be studied in an introductory course.
-In the context of binary classification, a labeled observation is a coordinate pair $(x, y)$ where $x\in\mathbb{R}^d$ is a vector of data associated with the observation (feature vector) and $y\in\{1, -1\}$ is the corresponding label of the observation.
-For example, $x_i$ might be a point in $\mathbb{R}^2$ representing the height and weight of the $i$th individual in a data set while the value of $y_i$ may correspond to the person's gender.
-The goal in classification is to accurately predict the label of a given observation using its feature vector.
-In this setting, the scenario of supervised learning involves using a training set of labeled observations $(x_i, y_i)$ to construct a prediction function $h$, one that takes a feature vector $\hat{x}$ and gives a prediction on the corresponding label $h(\hat{x})=\hat{y}$.
-It is common to use a separate set of test data to check the accuracy of the predictor, data that hasn't been used for training.
-There are many methods used to obtain a predictor for classification in supervised learning, and the class of ensemble methods concerns combining several predictors to obtain one.
+\section{Introduction} Binary and multiclass classification is a fundamental
+problem in machine learning and is commonly one of the first concepts to be
+studied in an introductory course. In the context of binary classification, a
+labeled observation is a coordinate pair $(x, y)$ where $x\in\mathbb{R}^d$ is a
+vector of data associated with the observation (feature vector) and $y\in\{1,
+-1\}$ is the corresponding label of the observation. For example, $x_i$ might
+be a point in $\mathbb{R}^2$ representing the height and weight of the $i$th
+individual in a data set while the value of $y_i$ may correspond to the
+person's gender. The goal in classification is to accurately predict the label
+of a given observation using its feature vector. In this setting, the scenario
+of supervised learning involves using a training set of labeled observations
+$(x_i, y_i)$ to construct a prediction function $h$, one that takes a feature
+vector $\hat{x}$ and gives a prediction on the corresponding label
+$h(\hat{x})=\hat{y}$. It is common to use a separate set of test data to check
+the accuracy of the predictor, data that hasn't been used for training. There
+are many methods used to obtain a predictor for classification in supervised
+learning, and the class of ensemble methods concerns combining several
+predictors to obtain one.
 
-Suppose now there exists an algorithm that, for any set of training data, is guaranteed to give you a predictor that has at least 50\% accuracy.
-Such an algorithm is called a weak learning algorithm since it yields a predictor that performs marginally better than guessing randomly.
-These prediction functions that a weak learning algorithm returns are called base classifiers, and a natural question arises about whether we can use a combination of weak learners to get a stronger one with much better accuracy and what this sort of ensemble method would look like.
-Ensemble methods like this are called boosting, and the algorithm AdaBoost introduced in \cite{no1} accomplishes this through successive rounds of training a weak learner to the data and subsequently re-weighting the training samples \cite{no1}.
+Suppose now there exists an algorithm that, for any set of training data, is
+guaranteed to give you a predictor that has at least 50\% accuracy. Such an
+algorithm is called a weak learning algorithm since it yields a predictor that
+performs marginally better than guessing randomly. These prediction functions
+that a weak learning algorithm returns are called base classifiers, and a
+natural question arises about whether we can use a combination of weak learners
+to get a stronger one with much better accuracy and what this sort of ensemble
+method would look like. Ensemble methods like this are called boosting, and the
+algorithm AdaBoost introduced in \cite{no1} accomplishes this through
+successive rounds of training a weak learner to the data and subsequently
+re-weighting the training samples \cite{no1}.
 
-The goal of this project is to implement a parallelized version of a commonly used weak learner, decision trees, to improve the running time of AdaBoost.
-We give a short overview of AdaBoost and decision trees in the next section, describe the implementation of the decision trees, the data sets used and the results t
+The goal of this project is to implement a parallelized version of a commonly
+used weak learner, decision trees, to improve the running time of AdaBoost. We
+give a short overview of AdaBoost and decision trees in the next section,
+describe the implementation of the decision trees, the data sets used and the
+results t
 
-\section{AdaBoost}
-The beauty of AdaBoost is in using a relatively inaccurate weak learning algorithm to produce a much more accurate prediction function.
-As mentioned, in each iteration of AdaBoost the weak learner selects a base classifier $h: \mathbb{R}^n\rightarrow\{1,-1\}$ from a set of functions $H$.
-The weak learner chooses in particular the base classifier  We give the algorithm below.\\
-\\
+\section{AdaBoost} The beauty of AdaBoost is in using a relatively inaccurate
+weak learning algorithm to produce a much more accurate prediction function. In
+each round of AdaBoost for $t$ from 1 to $T$, the weak learner selects a
+base classifier $h_t: \mathbb{R}^n\rightarrow\{1,-1\}$ from a set of functions
+$H$, in particular choosing the one with the least error
+$\epsilon_t=\sum_{i=1}^{n}D_t(i)I(h_t(x_i)\neq y_i)$ (Line 4) where $D_t(i)$ is
+the weight associated with the $i$th sample. The algorithm then reweights the
+training samples in Line 8 to place more emphasis on those that that the chosen
+classifier $h_t$ incorrectly predicted, so that in future rounds the base
+classifiers are tailored to correcting these mistakes. The predictor that is
+returned after all rounds is a linear combination of all $h_t$'s. The
+coefficient $\alpha_t$ gives more weight in the final predictor to those rounds
+where the classifier performed better. $Z_t$ is a normalization constant. We
+give the algorithm below.\\ \\
 {\sc AdaBoost}: $S = \{(x_i,y_i):i = 1\ldots n\}$
@@
 
+Note that everything except Line 4 can be done relatively quickly, so improving
+the speed of the weak learning algorithm will improve the speed of AdaBoost by
+a similar factor.
+
 \section{Decision Trees}
-Decision trees are a well-known method useful for classification which
-partitions the feature space into rectangular regions and gives each region the
-most common label among the training set points it contains. Each partition of
-the feature space is made along coordinate axes and done in a heirarchical way.
-Starting with the entire space, a line parallel to an axis divides it into two
-regions and then the same is done for these resulting regions. This process is
-repeated many times until the space is partitioned into several rectangles,
-like the example in Figure 1 (\cite{HTF}).
-Each split is made to separate points of the two classes distinct classes into different regions as best as possible.
-Thus the two resulting rectangles will have a greater {\it purity} in comparison to the parent node.
-There are a .
+Decision trees are a well-known method used in classification which partitions
+the feature space into rectangular regions and gives each region the most
+frequent label among the training set points it contains. In the most common
+decision tree algorithms each partition of the feature space is made along
+coordinate axes and done in a heirarchical way.  Starting with the entire
+space, a line parallel to an axis divides it into two regions and then the same
+is done for these resulting regions. This process is repeated many times until
+the space is partitioned into several rectangles, like the example in Figure 1
+(\cite{HTF}). Each split is made to separate points of the two distinct classes
+into different regions as best as possible. Thus the two resulting rectangles
+will have a greater {\it purity} in comparison to the parent node.
 
@@
 
- In that depiction, the lines drawn at $t_i$ represent the $i$th division of the space.
-Since each region is divided in two, this gives rise to a binary tree where each node represents a line drawn to partition the region existing before that node.
-The tree corresponding to Figure 1 is given in Figure 2 (\cite{HTF}).
-As depicted in the two figures, the splitting of each node is done in a heirarchical way so that .
-The feature space is partitioned along coordinate axes, which gives rise to decision.
-The most common way to partition the feature space is along coordinate axes.
+There are a few common ways to measure purity for a given region containing
+sample points. Gini Index is used here which can be easily substituted for
+others. For two classes the Gini Index of region $A$ is $I(A) = 2p(1-p)$ where
+$p$ is the probability of choosing a point in the first class from $A$, which
+translates to the probability of incorrectly guessing the class of a randomly
+chosen point from either class.
+
+The splitting of regions gives rise to the tree structure. At each node in the
+tree, the samples in the region corresponding to that node are ordered and
+counted to find in which dimension that region can be divided that gives the
+greatest increase in the purity of the branch. Figure 2 shows an example of
+this where at the top of the tree the decision to split on the first feature
+$X_1$ is best and the separation is made at $t_1$. In the first of the two
+resulting regions it is determined that the best split is made on $X_2$ at
+$t_2$, and in the other region the best separation is on $X_1$ at $t_3$. This
+splitting process will go on until a stopping criterion is met, either if there
+are too few sample points in a region or if no gain in purity can be made by
+splitting on any feature.
+
+%In Figure 1, the lines drawn at $t_i$ represent the division of the previous
+%region. Since each region is divided in two until a stopping criterion is met,
+%this gives rise to a binary tree where the action at each node is a line that
+%partitioning the region at that node. At each level of the tree the tree
+%corresponding to Figure 1 is given in Figure 2 (\cite{HTF}). As depicted in the
+%two figures, the splitting of each node is done in a heirarchical way so that .
+%The feature space is partitioned along coordinate axes, which gives rise to
+%decision. The most common way to partition the feature space is along
+%coordinate axes.
+
+Since the tree-building process is the most time-consuming portion of AdaBoost,
+our goal is to parallelize this. It seems natural to divide amongst processors
+the work of finding the dimension in the feature space that best splits the
+region and increases the purity for each node in the tree. The datasets we are
+handling have several dimensions, on the order of tens to hundreds, so each
+processor could be responsible for a group of dimensions in the feature space.
+After each processor determines the best dimension to split on amongst it's
+group, we can perform a reduction and determine the best split over all
+dimensions.
 

### 2015-08-21 paper.tex
paper changes

@@
 describe the implementation of the decision trees, the data sets used and the
-results t
+results comparing the serial and parallelized execution of the boosting algorithm.
 

### 2015-08-21 paper.tex
paper changes

@@
 be a point in $\mathbb{R}^2$ representing the height and weight of the $i$th
-individual in a data set while the value of $y_i$ may correspond to the
-person's gender. The goal in classification is to accurately predict the label
+individual in a data set while the value of $y_i$ may indicate the individual's
+gender. The goal in classification is to accurately predict the label
 of a given observation using its feature vector. In this setting, the scenario
@@
 vector $\hat{x}$ and gives a prediction on the corresponding label
-$h(\hat{x})=\hat{y}$. It is common to use a separate set of test data to check
-the accuracy of the predictor, data that hasn't been used for training. There
-are many methods used to obtain a predictor for classification in supervised
+$h(\hat{x})=\hat{y}$.
+To check the accuracy of the predictor it is common to use a separate set of test data, i.e.
+data that hasn't been used for training.
+There are many methods used to obtain a predictor for classification in supervised
 learning, and the class of ensemble methods concerns combining several
-predictors to obtain one.
+predictors together.
 
@@
 algorithm is called a weak learning algorithm since it yields a predictor that
-performs marginally better than guessing randomly. These prediction functions
-that a weak learning algorithm returns are called base classifiers, and a
-natural question arises about whether we can use a combination of weak learners
-to get a stronger one with much better accuracy and what this sort of ensemble
-method would look like. Ensemble methods like this are called boosting, and the
+performs marginally better than guessing randomly. The prediction functions
+returned by a weak learning algorithm are called base classifiers, and a
+natural question arises asking whether a combination of weak learners can be used
+to make a strong learner, or one that can achieve a certain level of accuracy.
+Ensemble methods like this are called boosting, and the
 algorithm AdaBoost introduced in \cite{no1} accomplishes this through
@@
 $\epsilon_t=\sum_{i=1}^{n}D_t(i)I(h_t(x_i)\neq y_i)$ (Line 4) where $D_t(i)$ is
-the weight associated with the $i$th sample. The algorithm then reweights the
+the weight associated with the $i$th sample. The algorithm then re-weights the
 training samples in Line 8 to place more emphasis on those that that the chosen

### 2015-08-21 paper.tex
paper changes

@@
 
-\section{Introduction} Binary and multiclass classification is a fundamental
+\section{Introduction}
+
+\indent Binary and multiclass classification is a fundamental
 problem in machine learning and is commonly one of the first concepts to be
@@
 
-The goal of this project is to implement a parallelized version of a commonly
-used weak learner, decision trees, to improve the running time of AdaBoost. We
+The goal of this project is to implement a parallelized version of decision trees, a commonly
+used weak learner, in order to improve the running time of AdaBoost. We
 give a short overview of AdaBoost and decision trees in the next section,
@@
 
-\section{AdaBoost} The beauty of AdaBoost is in using a relatively inaccurate
+\section{AdaBoost}
+
+\indent The beauty of AdaBoost is in using a relatively inaccurate
 weak learning algorithm to produce a much more accurate prediction function. In
@@
 $H$, in particular choosing the one with the least error
-$\epsilon_t=\sum_{i=1}^{n}D_t(i)I(h_t(x_i)\neq y_i)$ (Line 4) where $D_t(i)$ is
-the weight associated with the $i$th sample. The algorithm then re-weights the
-training samples in Line 8 to place more emphasis on those that that the chosen
-classifier $h_t$ incorrectly predicted, so that in future rounds the base
-classifiers are tailored to correcting these mistakes. The predictor that is
-returned after all rounds is a linear combination of all $h_t$'s. The
-coefficient $\alpha_t$ gives more weight in the final predictor to those rounds
-where the classifier performed better. $Z_t$ is a normalization constant. We
-give the algorithm below.\\ \\
+$\epsilon_t=\sum_{i=1}^{n}D_t(i)I(h_t(x_i)\neq y_i)$ (Line 4), where $D_t(i)$ is
+the weight associated with the $i$th sample.
+The coefficients $\alpha_t$ and $Z_t$ are calculated, where $\alpha_t$ is used in the ensemble prediction function and $Z_t$ is a normalization constant.
+The algorithm then updates the
+weights in Line 8 to place more emphasis on the samples that $h_t$ misclassified,
+so that in future rounds the base classifiers are chosen to correct these mistakes. The predictor that is
+returned in Line 9 after $T$ rounds is a linear combination of all $h_t$'s. Note the
+coefficient $\alpha_t$ gives more weight in the final predictor to the classifiers from rounds that fared better.
+We give the algorithm below.\\ \\
 {\sc AdaBoost}: $S = \{(x_i,y_i):i = 1\ldots n\}$
@@
 \item \quad \quad $D_{t+1}(i) = \frac{D_t(i)\exp(-\alpha_t y_i h_t(x_i))}{Z_t}$
-\item $g = \sum_{t=1}{T} \alpha_t h_t$
+\item $g = \sum_{t=1}^T \alpha_t h_t$
 \item {\bf return} $\mbox{sign}(g)$
@@
 \section{Decision Trees}
-Decision trees are a well-known method used in classification which partitions
+
+\indent Decision trees are a well-known method used in classification which partitions
 the feature space into rectangular regions and gives each region the most
@@
 processor could be responsible for a group of dimensions in the feature space.
-After each processor determines the best dimension to split on amongst it's
+After each processor determines the best dimension to split amongst its
 group, we can perform a reduction and determine the best split over all

### 2015-08-25 paper.tex
paper changes

--- 2015-08-21 paper.tex
@@
 processor could be responsible for a group of dimensions in the feature space.
-After each processor determines the best dimension to split on amongst it's
-group, we can perform a reduction and determine the best split over all
-dimensions.
+After each processor determines the best dimension to split on amongst those
+it's responsible for, we can perform a reduction and determine the best split
+over all dimensions. Any tree will split the feature space multiple times so
+there are several occasions for parallelism to speed up execution. In testing I
+commonly set the maximum depth for a tree to between two to five levels. Also
+the number of rounds of AdaBoost, and hence the number of trees constructed, is
+a parameter in the method allowing you to affect how often the algorithm drops
+into parallel structure.
+
+I've noticed with the MNIST I can achieve a speedup by a factor of around 4
+when using four processors, yet increasing beyond this does not give any
+noticeable gains.
+
+Executing this parallelized tree building
 

### 2015-08-25 paper.tex

--- 2015-08-21 paper.tex
@@
 processor could be responsible for a group of dimensions in the feature space.
-After each processor determines the best dimension to split amongst its
-group, we can perform a reduction and determine the best split over all
-dimensions.
+
+After each processor determines the best dimension to split on amongst those
+it's responsible for, we can perform a reduction and determine the best split
+over all dimensions. Any tree will split the feature space multiple times so
+there are several occasions for parallelism to speed up execution. In testing I
+commonly set the maximum depth for a tree to between two to five levels. Also
+the number of rounds of AdaBoost, and hence the number of trees constructed, is
+a parameter in the method allowing you to affect how often the algorithm drops
+into parallel structure.
+
+I've noticed with the MNIST I can achieve a speedup by a factor of around 4
+when using four processors, yet increasing beyond this does not give any
+noticeable gains.
+
+Executing this parallelized tree building
 

### 2015-09-04 paper.tex
paper additions

--- 2015-08-21 paper.tex
@@
 
-\indent Decision trees are a well-known method used in classification which partitions
-the feature space into rectangular regions and gives each region the most
-frequent label among the training set points it contains. In the most common
-decision tree algorithms each partition of the feature space is made along
-coordinate axes and done in a heirarchical way.  Starting with the entire
-space, a line parallel to an axis divides it into two regions and then the same
-is done for these resulting regions. This process is repeated many times until
-the space is partitioned into several rectangles, like the example in Figure 1
+\indent Decision trees are a well-known method used in classification which
+partitions the feature space into rectangular regions and applies to each
+region the most frequent label among the training set points it contains. In
+the most common decision tree algorithms each partition of the feature space is
+made along coordinate axes and is done in a heirarchical way. Initially, a line
+parallel to an axis divides the space into two regions, and then the same is
+done for these resulting regions. This process is repeated many times until the
+space is partitioned into several rectangles, like the example in Figure 1
 (\cite{HTF}). Each split is made to separate points of the two distinct classes
 into different regions as best as possible. Thus the two resulting rectangles
-will have a greater {\it purity} in comparison to the parent node.
+will have less {\it impurity} in comparison to the parent node.
 
-\begin{figure}[b!]
-\centering
-\begin{minipage}{0.45\textwidth}
-\centering
-\includegraphics[width=2.5in]{partition}
-\caption{space partitioned by a decision tree}
-\end{minipage}\hfill
-\begin{minipage}{0.45\textwidth}
-\centering
-\includegraphics[width=2.5in]{dectree}
-\caption{decision tree}
-\end{minipage}
-\end{figure}
-
-There are a few common ways to measure purity for a given region containing
-sample points. Gini Index is used here which can be easily substituted for
+There are a few common ways to measure impurity for a given region containing
+sample points. Gini index is used here which can be easily substituted for
 others. For two classes the Gini Index of region $A$ is $I(A) = 2p(1-p)$ where
@@
 counted to find in which dimension that region can be divided that gives the
-greatest increase in the purity of the branch. Figure 2 shows an example of
+greatest decrease in the impurity of the branch. Figure 2 shows an example of
 this where at the top of the tree the decision to split on the first feature
@@
 
+\begin{figure}[h!]
+\centering
+\begin{minipage}{0.45\textwidth}
+\centering
+\includegraphics[width=2.5in]{partition}
+\caption{space partitioned by a decision tree}
+\end{minipage}\hfill
+\begin{minipage}{0.45\textwidth}
+\centering
+\includegraphics[width=2.5in]{dectree}
+\caption{decision tree structure}
+\end{minipage}
+\end{figure}
+
+\section{Implementation}
+
 Since the tree-building process is the most time-consuming portion of AdaBoost,
@@
 processor could be responsible for a group of dimensions in the feature space.
-After each processor determines the best dimension to split amongst its
-group, we can perform a reduction and determine the best split over all
-dimensions.
+
+The AdaBoost algorithm is written in the main method of the program contained
+in {\tt paraboost.c}. This file also contains a few methods for utility, one
+that converts a decision tree data structure into a useable function and
+another that calculates weighted error. main() also performs load balancing by
+determing the features that each processor is responsible for.
+
+The methods that collect data from files on the hard disk and organize storage
+are defined in {\tt data.c}. These methods scrape floats from text and binary
+files and allocate space on the heap to store feature vectors and their labels.
+main() calls them early to load the training data in memory. There are a number
+of different sorting routines in {\tt sort.c}, all implementations of QuickSort
+or MergeSort that vary with the structure of the data types they sort. These
+methods are used to order data structures prior to the decision tree
+construction.
+
+The methods that implement decision tree operations are located in {\tt
+tree.c}. ParallelSplit() performs the parallel node splitting procedure for a
+given node in a decision tree, using the data points in that node to find the
+partition minimizing the impurity of the resulting child nodes. It calls the
+subroutine PodWBS() for each feature, which computes the impurity for every
+possible partition of the data over that feature dimension and returns an index
+corresponding to the best partition for that dimension and the impurity. Since
+this is done for every feature the processor is responsible for,
+ParallelSplit() then broadcasts which of its features is best and the minimum
+impurity to every processor. Thus an Allreduce is performed in order to
+determine which feature among all has the best partition achieving the minimum
+impurity. This determines the two branches of the decision tree, and the root
+processor informs the others which branch of the tree each observation goes to.
+The arrays of pointers to these observations are re-sorted to reflect this and
+ParallelSplit() is then called recursively on each of the two child nodes.
+After an end condition is met (depth in tree, small number of points
+in node, or very low impurity) the function returns. Thus this method is called for every node in the tree and performs noticeable computation for all nodes that aren't leaves.
+
+%After each processor determines the best dimension to split on amongst those
+%it's responsible for, we can perform a reduction and determine the best split
+%over all dimensions.
+
+Any tree will split the feature space multiple times so
+there are several occasions for parallelism to speed up execution. In testing I
+commonly set the maximum depth for a tree to between two to five levels. Also
+the number of rounds of AdaBoost, and hence the number of trees constructed, is
+a parameter in the method allowing you to affect how often the algorithm executes in parallel.
+
+\section{Data and Results}
+We employ the MNIST dataset to test our implementation.
+MNIST is a collection of 70,000 images of handwritten digits commonly used in image processing and pattern recognition research.
+The grayscale images are written in a binary file and consist of 28x28 arrays of integers ranging from 0 to 255 representing pixels varying from white to black.
+From this we separate out the set of 1's and 7's with the goal of performing binary classification between these two digits.
+We'll also trying to do the same thing with the digits 4 and 9.
+
+With the MNIST dataset, a speedup by a factor of roughly four can be achieved
+when using four processors, yet increasing beyond this does not give any
+noticeable gains.
 
@@
 
-%\bibitem{no4} Wiki
+%\bibitem{no4} Wiki AdaBoost
+
+%\bibitem{MNIST} Yann LeCun MNIST
+
 %\end{thebibliography}

### 2015-09-04 paraboost.c
changes to paraboost

@@ -6,8 +6,9 @@
 #include "mpi.h"
 
 
-/* paraboost.c implements a parallelized version of AdaBoost, kept in a
- * separate file to simplify compilation.
+/* paraboost.c implements a parallelized version of AdaBoost and contains
+ * the main section of the boosting algorithm.
+ * 
  */
 
 
@@ -48,7 +49,8 @@ double PError(Node *tree, double **data, Pod **base, int n) {
 
     for (i = 0; i < n; ++i) {
         if (WeakLearner(tree, data[i])*base[i]->label < 0)
-            error += base[i]->weight;
+            error += 1./n;
+            //error += base[i]->weight;
     }
 
     return error;
@@ -78,10 +80,10 @@ int main (int argc, char *argv[]) {
     }
 
 
-/* feature_list lists the features/indices that a process is responsible
- * for sortings and splitting. num_features is the length of feature_list.
- * Set feature_list and num_features for each process by dividing evenly
- * features over different processes.
+/* feature_list lists the features/columns that a process is responsible
+ * for sorting and splitting. num_features is the length of feature_list.
+ * Set feature_list and num_features for each process by dividing features
+ * evenly over different processes.
  */
     int num_features = (D-1)/p;
     int remainder = (D-1)%p;
@@ -100,6 +102,7 @@ int main (int argc, char *argv[]) {
     printf("Rank %d has feature %d through %d\n", rank, feature_list[0], feature_list[num_features-1]);
     
     //Copy data from files into memory
+    //TODO: Memory leak might be from array that MYDATA points to; none of MYDATA's elements, which are pointers, are freed
     double **MYDATA = ParHIGGS(feature_list, num_features);
 
     //return 0;
@@ -125,7 +128,7 @@ int main (int argc, char *argv[]) {
     //data is array of pointers to base, which get sorted by feature value
     Pod ***data = malloc(num_features*sizeof(Pod**));
     int feat;
-    for (feat = 0; feat < num_features; feat++){
+    for (feat = 0; feat < num_features; feat++) {
         data[feat] = malloc(n*sizeof(Pod*));
         for (i = 0; i < n; ++i)
             data[feat][i] = base[i];
@@ -134,9 +137,10 @@ int main (int argc, char *argv[]) {
     free(MYDATA);
 
     //Use ALLDATA to check training error and re-weight observations
[…]

### 2015-09-04 sort.c
changes to sort

@@ -1,5 +1,6 @@
 #include <stdlib.h>
 #include <stdio.h>
+#include <time.h>
 #include "header.h"
 
 /*
@@ -167,7 +168,7 @@ void PodSort(Pod **data, int first, int last, int feat) {
 
 /* PodSort is simply QuickSort but for data stored in Pods.
  *
- * data  = array containing sample data, labels, weights in Pod form to be sorted
+ * data  = pointer to array containing sample data, labels, weights in Pod form to be sorted
  * first = first unsorted row that Sort may move
  * last  = last unsorted row that Sort may move
  * feat  =  feature/index to sort on
@@ -192,7 +193,7 @@ void PodPartition(Pod **data, int first, int last, int feat, int ends[]) {
 
 /* PodPartition is simply Partition but for data stored in Pods.
  *
- * data  = array containing sample data, labels, weights in Pod form to be sorted
+ * data  = pointer to array containing sample data, labels, weights in Pod form to be sorted
  * first = first unsorted row that Sort may move
  * last  = last unsorted row that Sort may move
  * feat  =  feature/index to sort on
@@ -201,12 +202,18 @@ void PodPartition(Pod **data, int first, int last, int feat, int ends[]) {
  *
  */
 
+    srand(time(NULL));
+    int p_ind = rand() % (last-first+1) + first;
+    Pod *holder;
+    holder = data[last];
+    data[last] = data[p_ind];
+    data[p_ind] = holder;
+
     double pivot = data[last]->val[feat];
     int i = first;
     int j = last-1;
     int k = last;
     int l;
-    Pod *holder;
     while (i < j) {
         if (data[i]->val[feat] < pivot)
             ++i;

### 2015-09-04 tree.c
changes to tree

@@ -76,16 +76,16 @@ int BestSplit(double **data, int n, int first, int col, int pos, double *impurit
 }
 
 
-int WeightedBestSplit(double **data, int n, int first, int col, double pos, double tot, double *impurity) {
+int WeightedBestSplit(double **data, int n, int first, int feat, double pos, double tot, double *impurity) {
 
 /* Returns the row/index of the table with the least impurity after splitting
- * for fixed column/feature col. Partition rows up to and including that index
+ * for fixed column/feature feat. Partition rows up to and including that index
  * from everything afterwards.
  *
  * data     = array of data sorted on index a 
  * n        = length of table (# of rows/samples)
  * first    = first index in the node
- * col      = sorting/splitting feature/column of data
+ * feat      = sorting/splitting feature/column of data
  * pos      = weight of positive labels
  * tot      = total weight of all labels
  * impurity = pointer to save impurity after split
@@ -111,9 +111,9 @@ int WeightedBestSplit(double **data, int n, int first, int col, double pos, doub
     int i = 0;
     while (i < n) {
 
-        threshold = data[first+i][col];
+        threshold = data[first+i][feat];
 
-        while (i < n && (data[first+i][col] == threshold)) {
+        while (i < n && (data[first+i][feat] == threshold)) {
             if (data[first+i][D-1] > 0)
                 lpos += data[first+i][D];
 
@@ -216,6 +216,7 @@ void SplitNode(Node *node, double **data, int n, int first, int level) {
  *         column as the label (data[i][d-1]))
  * n     = length of table (# of rows/samples) on branch of node
  * first = first index of samples on branch of node
+ * level = the depth of node in the tree
  */
     
     timestamp_type sort_start, sort_stop, split_start, split_stop;
@@ -362,7 +363,7 @@ void ParallelSplit(Node *node, Pod ***data, int n, int first, int level, int ran
  * data for one feature plus the labels and a pointer to the sample ("pod")
  * with index to be used as a key.
  * 
- * Each processor has it's data pre-sorted after loading. 
+ * Each processor has its data sorted before ParallelSplit is called. 
  *
  */
 
@@ -442,7 +443,7 @@ void ParallelSplit(Node *node, Pod ***data, int n, int first, int level, int ran
     //get_timestamp(&split_stop);
     //split_time += timestamp_diff_in_seconds(split_start, split_stop);
 
-    //Save impurity and process rank to structure
+    //Save impurity and process rank to structure for MPI communication
     struct {
         double P; //impurity
         int R; //rank

### 2015-09-06 header.h
changes to header and paraboost

@@ -1,10 +1,12 @@
 #define GINI(pos, n) (2.*(pos)/(n)*((n)-(pos))/(n))
 #define ENTROPY(pos, n) ((pos)/(n)*log((pos)/(n)) + ((n)-(pos))/(n)*log(((n)-(pos))/(n)))
 #define WGINI(pos, tot) (2.*(pos)/(tot)*((tot)-(pos))/(tot))
-#define D 29
-#define N 1000000
+//#define D 29
+//#define N 1000000
+#define D 785
+#define N 13007
 
-//Note that D represents the number of features + 1 (for the label)
+//Note that D = number of features + 1 (for the label)
 
 struct TreeNode {
     struct TreeNode *parent;

### 2015-09-06 paraboost.c

@@ -49,8 +49,8 @@ double PError(Node *tree, double **data, Pod **base, int n) {
 
     for (i = 0; i < n; ++i) {
         if (WeakLearner(tree, data[i])*base[i]->label < 0)
-            error += 1./n;
-            //error += base[i]->weight;
+            //error += 1./n;
+            error += base[i]->weight;
     }
 
     return error;
@@ -103,7 +103,8 @@ int main (int argc, char *argv[]) {
     
     //Copy data from files into memory
     //TODO: Memory leak might be from array that MYDATA points to; none of MYDATA's elements, which are pointers, are freed
-    double **MYDATA = ParHIGGS(feature_list, num_features);
+    //double **MYDATA = ParHIGGS(feature_list, num_features);
+    double **MYDATA = ParMNIST17(feature_list, num_features);
 
     //return 0;
 
@@ -125,6 +126,7 @@ int main (int argc, char *argv[]) {
         base[i]->weight = 1./n; //weight
     }
 
+    printf("base weight = %f\n", base[0]->weight);
     //data is array of pointers to base, which get sorted by feature value
     Pod ***data = malloc(num_features*sizeof(Pod**));
     int feat;
@@ -138,9 +140,15 @@ int main (int argc, char *argv[]) {
 
     //Use ALLDATA to check training error and re-weight observations
     //TODO: Memory leak might be from array that ALLDATA points to
+    int *all_features = malloc((D-1)*sizeof(int));
+    for (i = 0; i < D-1; ++i)
+        all_features[i] = i;
     double **ALLDATA = NULL;
     if (rank == 0)
-        ALLDATA = ParHIGGS(feature_list, num_features);//MNIST17();
+        //ALLDATA = ParHIGGS(all_features, D-1);
+        ALLDATA = ParMNIST17(all_features, D-1);
+        
+    free(all_features);
 
     timestamp_type start, stop;
     int t;
@@ -230,10 +238,11 @@ int main (int argc, char *argv[]) {
         }
         
         MPI_Bcast(send_weights, n, MPI_DOUBLE, 0, MPI_COMM_WORLD);
-        printf("weights[0] = %f\n", send_weights[0]);
         for (i = 0; i < n; ++i)
             base[i]->weight = send_weights[i];
 
+        printf("weights[0] = %f\n", send_weights[0]);
+        printf("rank %d base weight = %f\n", rank, base[0]->weight);
         printf("t = %d: end\n", t);
     }
 

### 2015-09-08 paper.tex
paper changes

@@
 After an end condition is met (depth in tree, small number of points
-in node, or very low impurity) the function returns. Thus this method is called for every node in the tree and performs noticeable computation for all nodes that aren't leaves.
+in node, or very low impurity) the function returns. This method is called for every node in the tree and performs noticeable computation for all nodes that aren't leaves.
 
@@
 
-\section{Data and Results}
+\section{Data and Timings}
 We employ the MNIST dataset to test our implementation.
@@
 From this we separate out the set of 1's and 7's with the goal of performing binary classification between these two digits.
-We'll also trying to do the same thing with the digits 4 and 9.
+We'll also try this with the digits 4 and 9.
+This dataset can be obtained from \cite{MNIST}, and they are the four files referred to as the training set and test set images and labels.
 
-With the MNIST dataset, a speedup by a factor of roughly four can be achieved
-when using four processors, yet increasing beyond this does not give any
-noticeable gains.
+To test the parallel efficiency of this implementation, I ran the program on the host {\bf crunchy1} with 20 rounds of boosting with varying the number of processors as given in the following table.
+This gives the amount of time it took for the program to run as a whole in addition to the average time per processor spent constructing trees, since I wanted to measure the time spent in parallel without the extra overhead time performing serial tasks.
+
+%With the MNIST dataset, a speedup by a factor of roughly four can be achieved when using four processors, yet increasing beyond this does not give any noticeable gains.
+
+\section{Useage}
+I used MPICH for the MPI processor communications.
+To use the parallel boosting tool, include the file {\tt makefile} in the folder with the source code and compile it to create the executable {\tt pboost}.
+Using the command {\tt mpirun}, specify the number of processors and list the number of boosting rounds to perform after the executable.
+For instance, to run the executable with four processors for 50 rounds:
+
+\begin{center}
+{\tt mpirun -np 4 ./pboost 50}
+\end{center}
 

### 2015-09-08 paraboost.c
paraboost changes

@@ -151,6 +151,8 @@ int main (int argc, char *argv[]) {
     free(all_features);
 
     timestamp_type start, stop;
+    timestamp_type t_start, t_stop;
+    double t_elapsed = 0.;
     int t;
     int s;
     double e;
@@ -186,17 +188,19 @@ int main (int argc, char *argv[]) {
 
         //Sort data before tree-building
         for (feat = 0; feat < num_features; ++feat) {
-            printf("%d", feat);
+            //printf("%d", feat);
             PodSort(data[feat], 0, n-1, feat);
         }
-        printf("\n");
+        //printf("\n");
 
         printf("t = %d: done sorting\n", t);
 
+        get_timestamp(&t_start);
         //Building tree
         if (rank == 0)
             printf("t = %d: Building tree\n", t); 
         ParallelSplit(H[t], data, n, 0, 0, rank, num_features);
+        get_timestamp(&t_stop);
 
         if (rank == 0) {
             e = PError(H[t], ALLDATA, base, n);
@@ -241,16 +245,19 @@ int main (int argc, char *argv[]) {
         for (i = 0; i < n; ++i)
             base[i]->weight = send_weights[i];
 
-        printf("weights[0] = %f\n", send_weights[0]);
-        printf("rank %d base weight = %f\n", rank, base[0]->weight);
+        //printf("weights[0] = %f\n", send_weights[0]);
+        //printf("rank %d base weight = %f\n", rank, base[0]->weight);
         printf("t = %d: end\n", t);
+        t_elapsed += timestamp_diff_in_seconds(t_start, t_stop);
     }
 
     MPI_Barrier(MPI_COMM_WORLD);
     get_timestamp(&stop);
     
     double elapsed = timestamp_diff_in_seconds(start, stop);
+
     printf("Elapsed time: %f seconds\n", elapsed);
+    printf("Elapsed parallel tree-building time: %f seconds\n", t_elapsed);
 
     free(error);
     free(alpha);
@@ -277,18 +284,6 @@ int main (int argc, char *argv[]) {
 
     return 0;
 }
-
-
-
-
-//*/
-
-
-
-
-
-
-
 
 
 
